---
title: "Lab 5 - Data Wrangling"
author: "Arnav Bhatia"
format: html
link-citations: true
toc: false
---

# Learning goals

- Use the `merge()` function to join two datasets.
- Deal with missingness and impute data.
- Identify relevant observations using `quantile()`.

# Setup

```{r}
#| message: false
#| warning: false

# Install required packages if they are missing
required_packages <- c(
  "data.table",
  "dplyr",
  "leaflet",
  "R.utils"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load required packages
library(data.table)
library(dplyr)
library(leaflet)
library(R.utils)
```

```{r}
# Download the meteorological data
met_url <- paste0(
  "https://raw.githubusercontent.com/USCbiostats/",
  "data-science-data/master/02_met/met_all.gz"
)

met_file <- tempfile(fileext = ".gz")

download.file(
  url = met_url,
  destfile = met_file,
  mode = "wb"
)

# Read the meteorological data
met <- fread(met_file)
met <- as.data.frame(met)

# Download the NOAA station data
stations <- read.csv(
  "https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv"
)

# Convert the station identification number to integer
stations$USAF <- as.integer(stations$USAF)

# Replace invalid and blank values with NA
stations$USAF[stations$USAF == 999999] <- NA
stations$CTRY[stations$CTRY == ""] <- NA
stations$STATE[stations$STATE == ""] <- NA

# Select the relevant columns and retain unique records
stations <- unique(
  stations[, c("USAF", "CTRY", "STATE")]
)

# Remove stations without a valid USAF number
stations <- stations[!is.na(stations$USAF), ]

# Remove duplicate USAF station numbers
stations <- stations[
  !duplicated(stations$USAF),
]

# Merge the meteorological and station datasets
met <- merge(
  x = met,
  y = stations,
  by.x = "USAFID",
  by.y = "USAF",
  all.x = TRUE,
  all.y = FALSE
)

# Keep U.S. observations with an identified state
met_us <- met |>
  filter(
    CTRY == "US",
    !is.na(STATE)
  )

# Examine the merged dataset
dim(met_us)
head(met_us)
```

# Question 1: Representative station for the United States

I first calculated the mean temperature, wind speed, and atmospheric pressure
for every weather station. I then used `quantile()` to calculate the median of
each station-level measurement. The representative station was defined as the
station whose mean value was closest to the corresponding national median.

```{r}
# Create one row per weather station
station_summary <- met_us |>
  group_by(USAFID, STATE) |>
  summarise(
    lat = first(
      lat[!is.na(lat)],
      default = NA_real_
    ),
    lon = first(
      lon[!is.na(lon)],
      default = NA_real_
    ),
    mean_temp = ifelse(
      all(is.na(temp)),
      NA_real_,
      mean(temp, na.rm = TRUE)
    ),
    mean_wind = ifelse(
      all(is.na(wind.sp)),
      NA_real_,
      mean(wind.sp, na.rm = TRUE)
    ),
    mean_pressure = ifelse(
      all(is.na(atm.press)),
      NA_real_,
      mean(atm.press, na.rm = TRUE)
    ),
    .groups = "drop"
  )

# Calculate the national median of each station-level measurement
median_temp <- quantile(
  station_summary$mean_temp,
  probs = 0.50,
  na.rm = TRUE
)

median_wind <- quantile(
  station_summary$mean_wind,
  probs = 0.50,
  na.rm = TRUE
)

median_pressure <- quantile(
  station_summary$mean_pressure,
  probs = 0.50,
  na.rm = TRUE
)

# Create a table containing the national medians
national_medians <- data.frame(
  measurement = c(
    "Temperature",
    "Wind speed",
    "Atmospheric pressure"
  ),
  median = c(
    median_temp,
    median_wind,
    median_pressure
  )
)

national_medians
```

```{r}
# Find the station closest to the median temperature
temp_station <- station_summary |>
  filter(!is.na(mean_temp)) |>
  mutate(
    distance_from_median =
      abs(mean_temp - median_temp)
  ) |>
  slice_min(
    order_by = distance_from_median,
    n = 1,
    with_ties = FALSE
  ) |>
  select(
    USAFID,
    STATE,
    lat,
    lon,
    mean_temp,
    distance_from_median
  )

# Find the station closest to the median wind speed
wind_station <- station_summary |>
  filter(!is.na(mean_wind)) |>
  mutate(
    distance_from_median =
      abs(mean_wind - median_wind)
  ) |>
  slice_min(
    order_by = distance_from_median,
    n = 1,
    with_ties = FALSE
  ) |>
  select(
    USAFID,
    STATE,
    lat,
    lon,
    mean_wind,
    distance_from_median
  )

# Find the station closest to the median atmospheric pressure
pressure_station <- station_summary |>
  filter(!is.na(mean_pressure)) |>
  mutate(
    distance_from_median =
      abs(mean_pressure - median_pressure)
  ) |>
  slice_min(
    order_by = distance_from_median,
    n = 1,
    with_ties = FALSE
  ) |>
  select(
    USAFID,
    STATE,
    lat,
    lon,
    mean_pressure,
    distance_from_median
  )

# Display the three representative stations
temp_station
wind_station
pressure_station
```

```{r}
# Check whether the three measurements identify the same station
representative_station_ids <- c(
  temp_station$USAFID,
  wind_station$USAFID,
  pressure_station$USAFID
)

same_station <- length(
  unique(representative_station_ids)
) == 1

same_station
```

```{r}
# Automatically generate a written conclusion
if (same_station) {
  cat(
    "The three representative stations coincide. ",
    "The same station was closest to the median ",
    "temperature, wind speed, and atmospheric pressure."
  )
} else {
  cat(
    "The three representative stations do not coincide. ",
    "Different stations were closest to the median ",
    "temperature, wind speed, and atmospheric pressure."
  )
}
```

# Question 2: Representative station coordinates by state

For each state, I calculated the median latitude and median longitude among
its weather stations. I used the station-level dataset so stations with more
weather observations would not receive greater weight.

```{r}
state_median_coordinates <- station_summary |>
  filter(
    !is.na(STATE),
    !is.na(lat),
    !is.na(lon)
  ) |>
  group_by(STATE) |>
  summarise(
    median_lat = quantile(
      lat,
      probs = 0.50,
      na.rm = TRUE
    ),
    median_lon = quantile(
      lon,
      probs = 0.50,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(STATE)

state_median_coordinates
```

# Question 3: In the middle?

For each state, I calculated the mean latitude and longitude of its weather
stations. I then calculated the distance between each station and its state's
mean coordinates. The station with the smallest distance was selected as the
station closest to the middle of that state's available stations.

```{r}
# Calculate the mean latitude and longitude for each state
state_mean_coordinates <- station_summary |>
  filter(
    !is.na(STATE),
    !is.na(lat),
    !is.na(lon)
  ) |>
  group_by(STATE) |>
  summarise(
    mean_lat = mean(
      lat,
      na.rm = TRUE
    ),
    mean_lon = mean(
      lon,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

state_mean_coordinates
```

```{r}
# Add the state mean coordinates to every station
station_distances <- station_summary |>
  filter(
    !is.na(STATE),
    !is.na(lat),
    !is.na(lon)
  ) |>
  left_join(
    state_mean_coordinates,
    by = "STATE"
  ) |>
  mutate(
    distance_from_mean = sqrt(
      (lat - mean_lat)^2 +
      (lon - mean_lon)^2
    )
  )

# Select the closest station in each state
closest_state_stations <- station_distances |>
  group_by(STATE) |>
  slice_min(
    order_by = distance_from_mean,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup() |>
  arrange(STATE)

# Display the selected stations
closest_state_stations |>
  select(
    STATE,
    USAFID,
    lat,
    lon,
    mean_lat,
    mean_lon,
    distance_from_mean
  )
```

```{r}
# Create labels for the median-coordinate points
state_median_coordinates <- state_median_coordinates |>
  mutate(
    marker_label = paste0(
      STATE,
      ": median coordinates"
    )
  )

# Create labels for the selected stations
closest_state_stations <- closest_state_stations |>
  mutate(
    marker_label = paste0(
      STATE,
      ": closest station ",
      USAFID
    )
  )

# Create the interactive map
leaflet() |>
  addTiles(
    urlTemplate = "https://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png",
    attribution = "&copy; OpenStreetMap contributors"
  ) |>
  addCircleMarkers(
    data = state_median_coordinates,
    lng = ~median_lon,
    lat = ~median_lat,
    radius = 5,
    color = "blue",
    fillColor = "blue",
    fillOpacity = 0.75,
    stroke = FALSE,
    label = ~marker_label,
    group = "Median state coordinates"
  ) |>
  addCircleMarkers(
    data = closest_state_stations,
    lng = ~lon,
    lat = ~lat,
    radius = 5,
    color = "red",
    fillColor = "red",
    fillOpacity = 0.75,
    stroke = FALSE,
    label = ~marker_label,
    group = "Stations closest to mean coordinates"
  ) |>
  addLayersControl(
    overlayGroups = c(
      "Median state coordinates",
      "Stations closest to mean coordinates"
    ),
    options = layersControlOptions(
      collapsed = FALSE
    )
  ) |>
  addLegend(
    position = "bottomright",
    colors = c("blue", "red"),
    labels = c(
      "Median state coordinates",
      "Station closest to mean coordinates"
    ),
    opacity = 0.75,
    title = "Point type"
  )
```

The blue points represent the median latitude and longitude calculated in
Question 2. The red points represent the actual weather stations closest to
the mean latitude and longitude of the stations within each state.

# Question 4: Means of means

I first calculated the mean temperature for every state. I classified each
state as Low, Mid, or High according to its mean temperature. I then assigned
the state-level category to every observation and summarized the observations
within each category.

An NA entry was defined as a record with a missing value for at least one of
the three weather measurements: temperature, wind speed, or atmospheric
pressure.

```{r}
# Calculate the mean temperature for each state
state_temperature <- met_us |>
  filter(!is.na(STATE)) |>
  group_by(STATE) |>
  summarise(
    state_mean_temp = ifelse(
      all(is.na(temp)),
      NA_real_,
      mean(temp, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  mutate(
    temperature_category = case_when(
      state_mean_temp < 20 ~ "Low",
      state_mean_temp >= 20 &
        state_mean_temp < 25 ~ "Mid",
      state_mean_temp >= 25 ~ "High",
      TRUE ~ NA_character_
    ),
    temperature_category = factor(
      temperature_category,
      levels = c(
        "Low",
        "Mid",
        "High"
      )
    )
  ) |>
  arrange(
    temperature_category,
    STATE
  )

# Display the states and their classifications
state_temperature
```

```{r}
# Add the state temperature category to each observation
met_classified <- met_us |>
  left_join(
    state_temperature,
    by = "STATE"
  ) |>
  mutate(
    missing_weather_entry =
      is.na(temp) |
      is.na(wind.sp) |
      is.na(atm.press)
  )

# Generate the final summary table
category_summary <- met_classified |>
  filter(
    !is.na(temperature_category)
  ) |>
  group_by(
    temperature_category
  ) |>
  summarise(
    number_of_states =
      n_distinct(STATE),
    number_of_entries =
      n(),
    number_of_NA_entries =
      sum(missing_weather_entry),
    number_of_stations =
      n_distinct(USAFID),
    mean_temperature =
      mean(temp, na.rm = TRUE),
    mean_wind_speed =
      mean(wind.sp, na.rm = TRUE),
    mean_atmospheric_pressure =
      mean(atm.press, na.rm = TRUE),
    .groups = "drop"
  )

category_summary
```

The table reports the number of states, total number of observations, number
of observations containing missing weather data, number of unique stations,
mean temperature, mean wind speed, and mean atmospheric pressure within the
Low, Mid, and High state-temperature categories.
