---
title: "Assignment 1 - Exploratory Data Analysis"
author: "Arnav Bhatia"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
link-citations: true
execute:
  warning: false
  message: false
---

# Introduction

This exploratory data analysis examines whether daily concentrations of fine
particulate matter (PM$_{2.5}$) decreased in California between 2002 and 2022.
The data were downloaded from the U.S. Environmental Protection Agency's
[EPA Air Quality Data website](https://www.epa.gov/outdoor-air-quality-data/download-daily-data).
The analysis examines the data at the state, county, and Los Angeles County
monitoring-site levels.

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(janitor)
library(maps)
library(knitr)

theme_set(theme_minimal())

options(
  dplyr.summarise.inform = FALSE,
  scipen = 999
)
```

# 1. Initial Data Inspection

## Read the datasets

The CSV files are stored outside the GitHub repository so the raw EPA data will
not be uploaded to GitHub. On my computer, the files are stored in the Downloads
folder.

```{r}
# Change this line if the CSV files are stored somewhere else.
data_dir <- "~/Downloads"

pm2002_raw <- read_csv(
  file.path(data_dir, "ad_viz_plotval_data (1).csv"),
  show_col_types = FALSE
)

pm2022_raw <- read_csv(
  file.path(data_dir, "ad_viz_plotval_data.csv"),
  show_col_types = FALSE
)
```

## Dimensions

```{r}
data_dimensions <- tibble(
  year = c(2002, 2022),
  rows = c(nrow(pm2002_raw), nrow(pm2022_raw)),
  columns = c(ncol(pm2002_raw), ncol(pm2022_raw))
)

kable(
  data_dimensions,
  caption = "Dimensions of the two EPA datasets"
)
```

The 2002 dataset contains `r format(nrow(pm2002_raw), big.mark = ",")`
observations and `r ncol(pm2002_raw)` variables. The 2022 dataset contains
`r format(nrow(pm2022_raw), big.mark = ",")` observations and
`r ncol(pm2022_raw)` variables. Therefore, substantially more monitoring
observations were available in 2022.

## Headers and footers

```{r}
#| results: hold

head(pm2002_raw)
tail(pm2002_raw)

head(pm2022_raw)
tail(pm2022_raw)
```

The headers and footers confirm that the 2002 file begins and ends with dates
from 2002, while the 2022 file begins and ends with dates from 2022. Both files
contain observations from California and have the same basic structure.

## Variable names

```{r}
names(pm2002_raw)
names(pm2022_raw)
```

Both datasets have the same variable names. Important variables include
`Date`, `Site ID`, `Daily Mean PM2.5 Concentration`, `Local Site Name`,
`County`, `Site Latitude`, and `Site Longitude`.

## Variable types

```{r}
glimpse(pm2002_raw)
glimpse(pm2022_raw)
```

Most identification and location variables were read as character variables,
while PM$_{2.5}$, latitude, and longitude were read as numeric variables.
The `Date` variable was initially read as a character variable and will be
converted to a proper date in the next section.

## Distribution of PM2.5

```{r}
pm_distribution_raw <- bind_rows(
  pm2002_raw |> mutate(dataset_year = 2002),
  pm2022_raw |> mutate(dataset_year = 2022)
) |>
  transmute(
    year = factor(dataset_year),
    pm25 = `Daily Mean PM2.5 Concentration`
  )

pm_distribution_raw |>
  group_by(year) |>
  summarise(
    n = n(),
    missing = sum(is.na(pm25)),
    minimum = min(pm25, na.rm = TRUE),
    first_quartile = quantile(pm25, 0.25, na.rm = TRUE),
    median = median(pm25, na.rm = TRUE),
    mean = mean(pm25, na.rm = TRUE),
    third_quartile = quantile(pm25, 0.75, na.rm = TRUE),
    maximum = max(pm25, na.rm = TRUE)
  ) |>
  kable(
    digits = 2,
    caption = "Initial summary of daily PM2.5 concentrations"
  )
```

```{r}
ggplot(
  pm_distribution_raw,
  aes(x = pm25, fill = year)
) +
  geom_histogram(
    bins = 50,
    color = "white",
    alpha = 0.75
  ) +
  facet_wrap(
    vars(year),
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = c("2002" = "#2C7BB6", "2022" = "#D7191C")
  ) +
  labs(
    title = "Distribution of Daily PM2.5 Concentrations",
    subtitle = "California EPA monitoring observations, 2002 and 2022",
    x = expression("Daily mean PM"[2.5] * " concentration (" * mu * "g/m"^3 * ")"),
    y = "Number of observations",
    fill = "Year"
  ) +
  guides(fill = "none")
```

The distributions are right-skewed in both years, meaning that most daily
concentrations are relatively low but a smaller number of days have very high
concentrations. The 2022 data include negative values, which are not physically
plausible concentrations and will be addressed during the data-quality check.
The high positive observations may reflect genuine extreme pollution episodes,
including wildfire smoke, so they are retained unless there is additional
evidence that they are errors.

# 2. Combine and Clean the Data

The two datasets were combined using `bind_rows()`. The variable names were
standardized with `clean_names()`, and the key variables were given shorter
names. The year identifier was derived from the date rather than from the
dataset name.

```{r}
pm_combined <- bind_rows(pm2002_raw, pm2022_raw) |>
  clean_names() |>
  transmute(
    date = mdy(date),
    year = year(date),
    pm25 = daily_mean_pm2_5_concentration,
    site_id = as.character(site_id),
    poc = as.character(poc),
    site_name = local_site_name,
    county,
    latitude = site_latitude,
    longitude = site_longitude,
    parameter_code = as.character(aqs_parameter_code),
    parameter_description = aqs_parameter_description,
    method_code = as.character(method_code),
    method_description
  )
```

```{r}
glimpse(pm_combined)
```

```{r}
pm_combined |>
  count(year, name = "number_of_rows") |>
  kable(
    caption = "Number of observations after combining the datasets"
  )
```

The combined dataset contains observations from both years. The renamed
variables make the subsequent code shorter and easier to interpret.

Some sites have more than one measurement for the same date because the EPA
files include different parameter codes, instruments, or pollutant occurrence
codes. For the state, county, and site comparisons, these repeated measurements
are averaged to create one PM$_{2.5}$ value per site per day. This prevents
site-days with several instruments from receiving more weight than site-days
with only one measurement.

# 3. Spatial Distribution of Monitoring Sites

First, one record was retained for every unique monitoring site in each year.

```{r}
monitoring_sites <- pm_combined |>
  distinct(
    year,
    site_id,
    site_name,
    county,
    latitude,
    longitude
  )
```

```{r}
monitoring_sites |>
  count(year, name = "number_of_sites") |>
  kable(
    caption = "Number of California monitoring sites by year"
  )
```

```{r}
california_map <- map_data("state") |>
  filter(region == "california")

ggplot() +
  geom_polygon(
    data = california_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95",
    color = "gray40",
    linewidth = 0.4
  ) +
  geom_point(
    data = monitoring_sites,
    aes(
      x = longitude,
      y = latitude,
      color = factor(year)
    ),
    alpha = 0.70,
    size = 1.8
  ) +
  coord_quickmap() +
  scale_color_manual(
    values = c("2002" = "#2C7BB6", "2022" = "#D7191C")
  ) +
  labs(
    title = "California PM2.5 Monitoring Sites",
    subtitle = "Site locations in 2002 and 2022",
    x = "Longitude",
    y = "Latitude",
    color = "Year"
  )
```

There were `r n_distinct(filter(monitoring_sites, year == 2002)$site_id)`
monitoring sites in 2002 and
`r n_distinct(filter(monitoring_sites, year == 2022)$site_id)` sites in 2022.
In both years, monitoring sites were concentrated around major population
centers, including the Los Angeles region, the San Francisco Bay Area, the
Central Valley, and the San Diego region. The 2022 network included more sites
and provided broader coverage than the 2002 network. Nevertheless, the sites
were not distributed uniformly across California; rural and sparsely populated
areas generally had fewer monitors.

# 4. Missing and Implausible PM2.5 Values

A negative PM$_{2.5}$ concentration was classified as implausible because a
physical concentration cannot be below zero. Zero was retained as a possible
very low measurement. High positive values were also retained because major
air-pollution and wildfire events can produce extremely high concentrations.

```{r}
data_quality <- pm_combined |>
  mutate(
    missing_pm25 = is.na(pm25),
    implausible_pm25 = !is.na(pm25) & pm25 < 0,
    missing_or_implausible = missing_pm25 | implausible_pm25
  ) |>
  group_by(year) |>
  summarise(
    total_observations = n(),
    missing_n = sum(missing_pm25),
    missing_proportion = mean(missing_pm25),
    implausible_n = sum(implausible_pm25),
    implausible_proportion = mean(implausible_pm25),
    affected_n = sum(missing_or_implausible),
    affected_proportion = mean(missing_or_implausible)
  )

data_quality |>
  mutate(
    missing_proportion = scales::percent(
      missing_proportion,
      accuracy = 0.01
    ),
    implausible_proportion = scales::percent(
      implausible_proportion,
      accuracy = 0.01
    ),
    affected_proportion = scales::percent(
      affected_proportion,
      accuracy = 0.01
    )
  ) |>
  kable(
    caption = "Missing and implausible PM2.5 observations by year"
  )
```

There were no missing PM$_{2.5}$ values in either dataset. There were no
negative values in 2002, but the 2022 dataset contained
`r sum(pm_combined$year == 2022 & pm_combined$pm25 < 0, na.rm = TRUE)`
negative values. Thus, the identified implausible-value problem was limited to
2022, although it affected only a small proportion of that year's observations.

The implausible values were changed to missing before conducting the main
comparisons.

```{r}
pm_clean <- pm_combined |>
  mutate(
    pm25_clean = if_else(pm25 < 0, NA_real_, pm25)
  )
```

A site-day dataset was then created by averaging valid measurements recorded at
the same site on the same date.

```{r}
site_day <- pm_clean |>
  group_by(
    year,
    date,
    site_id,
    site_name,
    county,
    latitude,
    longitude
  ) |>
  summarise(
    pm25 = if (
      all(is.na(pm25_clean))
    ) {
      NA_real_
    } else {
      mean(pm25_clean, na.rm = TRUE)
    },
    number_of_measurements = n(),
    .groups = "drop"
  )
```

# 5. PM2.5 at Three Spatial Levels

## Level 1: State

The statewide comparison uses one observation per monitoring site per day.

```{r}
state_summary <- site_day |>
  group_by(year) |>
  summarise(
    site_days = n(),
    valid_site_days = sum(!is.na(pm25)),
    mean_pm25 = mean(pm25, na.rm = TRUE),
    standard_deviation = sd(pm25, na.rm = TRUE),
    minimum = min(pm25, na.rm = TRUE),
    first_quartile = quantile(pm25, 0.25, na.rm = TRUE),
    median_pm25 = median(pm25, na.rm = TRUE),
    third_quartile = quantile(pm25, 0.75, na.rm = TRUE),
    maximum = max(pm25, na.rm = TRUE)
  )

state_summary |>
  kable(
    digits = 2,
    caption = "Statewide summary of daily PM2.5 concentrations"
  )
```

```{r}
ggplot(
  site_day,
  aes(
    x = factor(year),
    y = pm25,
    fill = factor(year)
  )
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.65,
    na.rm = TRUE
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    fill = "white",
    na.rm = TRUE
  ) +
  coord_cartesian(ylim = c(0, 60)) +
  scale_fill_manual(
    values = c("2002" = "#2C7BB6", "2022" = "#D7191C")
  ) +
  labs(
    title = "Statewide Distribution of Daily PM2.5",
    subtitle = "The y-axis is limited to 60 µg/m³ so the central distributions are visible",
    x = "Year",
    y = expression("Daily mean PM"[2.5] * " (" * mu * "g/m"^3 * ")")
  ) +
  guides(fill = "none")
```

The statewide mean decreased from
`r round(state_summary$mean_pm25[state_summary$year == 2002], 1)`
µg/m³ in 2002 to
`r round(state_summary$mean_pm25[state_summary$year == 2022], 1)`
µg/m³ in 2022. The median decreased from
`r round(state_summary$median_pm25[state_summary$year == 2002], 1)`
µg/m³ to
`r round(state_summary$median_pm25[state_summary$year == 2022], 1)`
µg/m³. The lower mean and median in 2022 provide descriptive evidence that
typical daily PM$_{2.5}$ concentrations decreased statewide. However, both
years contain high-pollution observations, and the maximum alone should not be
used to describe the typical exposure level.

## Level 2: County

County-level summary statistics were calculated separately for every county and
year.

```{r}
county_summary <- site_day |>
  group_by(county, year) |>
  summarise(
    number_of_sites = n_distinct(site_id),
    site_days = n(),
    valid_site_days = sum(!is.na(pm25)),
    mean_pm25 = mean(pm25, na.rm = TRUE),
    median_pm25 = median(pm25, na.rm = TRUE),
    standard_deviation = sd(pm25, na.rm = TRUE),
    first_quartile = quantile(pm25, 0.25, na.rm = TRUE),
    third_quartile = quantile(pm25, 0.75, na.rm = TRUE)
  ) |>
  arrange(county, year)

county_summary |>
  kable(
    digits = 2,
    caption = "County-level PM2.5 summary statistics"
  )
```

```{r}
county_order <- county_summary |>
  group_by(county) |>
  summarise(overall_median = median(median_pm25, na.rm = TRUE)) |>
  arrange(overall_median) |>
  pull(county)

county_summary <- county_summary |>
  mutate(
    county = factor(county, levels = county_order)
  )

ggplot(
  county_summary,
  aes(
    x = median_pm25,
    y = county,
    color = factor(year),
    group = county
  )
) +
  geom_line(
    color = "gray70",
    linewidth = 0.6
  ) +
  geom_point(size = 2.2) +
  scale_color_manual(
    values = c("2002" = "#2C7BB6", "2022" = "#D7191C")
  ) +
  labs(
    title = "Median Daily PM2.5 by California County",
    subtitle = "Counties are ordered by their median concentration",
    x = expression("Median daily PM"[2.5] * " (" * mu * "g/m"^3 * ")"),
    y = NULL,
    color = "Year"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 7)
  )
```

The plot shows that the amount of change varied across counties. For counties
with data in both years, a 2022 point located to the left of its 2002 point
indicates a reduction in the median daily concentration. Many counties show
lower median PM$_{2.5}$ concentrations in 2022, although the magnitude of the
decrease is not identical across the state. Counties represented in only one
year cannot be used for a direct temporal comparison.

The number of counties represented also changed from
`r n_distinct(filter(site_day, year == 2002)$county)` in 2002 to
`r n_distinct(filter(site_day, year == 2022)$county)` in 2022. Therefore,
county comparisons should be interpreted with awareness that the monitoring
network changed over time.

## Level 3: Monitoring Sites in Los Angeles County

The final analysis was restricted to monitoring sites located in Los Angeles
County.

```{r}
la_site_day <- site_day |>
  filter(county == "Los Angeles") |>
  mutate(
    site_label = if_else(
      is.na(site_name) | site_name == "",
      site_id,
      paste0(site_name, " (", site_id, ")")
    )
  )
```

```{r}
la_site_summary <- la_site_day |>
  group_by(site_id, site_name, year) |>
  summarise(
    site_days = n(),
    valid_site_days = sum(!is.na(pm25)),
    mean_pm25 = mean(pm25, na.rm = TRUE),
    median_pm25 = median(pm25, na.rm = TRUE),
    standard_deviation = sd(pm25, na.rm = TRUE),
    minimum = min(pm25, na.rm = TRUE),
    maximum = max(pm25, na.rm = TRUE)
  ) |>
  arrange(site_name, site_id, year)

la_site_summary |>
  kable(
    digits = 2,
    caption = "PM2.5 summary statistics for Los Angeles County monitoring sites"
  )
```

```{r}
ggplot(
  la_site_day,
  aes(
    x = factor(year),
    y = pm25,
    fill = factor(year)
  )
) +
  geom_boxplot(
    outlier.alpha = 0.20,
    na.rm = TRUE
  ) +
  facet_wrap(
    vars(site_label),
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = c("2002" = "#2C7BB6", "2022" = "#D7191C")
  ) +
  labs(
    title = "Daily PM2.5 at Los Angeles County Monitoring Sites",
    subtitle = "Each panel represents one EPA monitoring site",
    x = "Year",
    y = expression("Daily mean PM"[2.5] * " (" * mu * "g/m"^3 * ")"),
    fill = "Year"
  ) +
  theme(
    strip.text = element_text(size = 7),
    legend.position = "bottom"
  )
```

The Los Angeles County results demonstrate that pollution levels and temporal
changes were not uniform across monitoring sites. Sites with observations in
both years generally allow the clearest direct comparison. Several sites show
lower central concentrations in 2022, but some sites were added, removed, or
relocated between the two years. Consequently, changes in the set of active
monitors must be considered when interpreting the Los Angeles County results.

# Conclusion

The exploratory results indicate that typical daily PM$_{2.5}$ concentrations
in California were lower in 2022 than in 2002. This pattern is visible in the
statewide mean and median and across many counties and Los Angeles County
monitoring sites. However, the monitoring network expanded and changed during
the 20-year period, so the two datasets do not represent exactly the same set
of locations.

The analysis also identified negative PM$_{2.5}$ observations in 2022. These
values were classified as implausible and excluded from the concentration
summaries. High positive values were retained because they may represent real
extreme pollution events. Overall, the results support the conclusion that
typical PM$_{2.5}$ concentrations decreased, while also showing meaningful
spatial variation and occasional extreme pollution episodes.
